---
title: "MicrobiomeExperiment/phyloseq low-level data translation"
author: "Domenick Braccia"
date: "01/17/2018"
output:
  BiocStyle::html_document:
    fig_height: 7
    fig_width: 10
    toc: yes
    toc_depth: 2
    number_sections: true
---
<!--
%% \VignetteEngine{knitr::rmarkdown}
%% \VignetteIndexEntry{MicrobiomeExperiment basics vignette}
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Introduction
This vignette is to show how to use base, SummarizedExperiment or metagenomeFeatures operations to perform standard phyloseq functions detailed in [this github issue](https://github.com/HCBravoLab/MicrobiomeExperiment/issues/14).

The dataset being used throughout this file is one of the examples from the [metagenomeFeatures package](http://bioconductor.org/packages/release/bioc/html/metagenomeFeatures.html) named `mock_mgF`.

# Making MicrobiomeExperiment object
```{r, warning=FALSE, message=FALSE}
# PACKAGES
library(S4Vectors)
#library(metagenomeSeq) # only needed if mouseData is being used
library(metagenomeFeatures)
library(TreeSummarizedExperiment)
library(MicrobiomeExperiment)

# from metagenomeFeatures
data(mock_mgF)

# making colData
pd <- DataFrame(a=letters[1:4], b=1:4)
numcounts <- nrow(mock_mgF) * 4

# making assay data
counts <- matrix(sample(1:1000,numcounts,replace=TRUE), nr=nrow(mock_mgF), nc=4)

# making rowData
taxa_table <- as.data.frame(DataFrame(mock_mgF@listData))
taxa_table <- taxa_table[, c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species", "OTU")]
#tree_index <- TreeIndex(taxa_table, feature_order = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species", "OTU")) #already done in MicrobiomeFeatures function

# MicrobiomeExperiment object creation
mbF <- MicrobiomeFeatures(taxa_table, phyloTree = mock_mgF@refDbTree, seq = mock_mgF@refDbSeq, metadata = mock_mgF@metadata)
MBExp <- MicrobiomeExperiment(assays=SimpleList(counts=counts),
                     rowData = mbF,
                     colData = pd)

# ## using mouseData ##
# library(metagenomeSeq)
# data(mouseData)
# 
# counts <- MRcounts(mouseData)
# taxTab <- fData(mouseData)
# 
# tree <- TreeIndex(taxTab)
# mbiome <- TreeSummarizedExperiment(SimpleList(counts=counts),
#                                    rowData=tree)
# #####################

```
# Making corresponding phyloseq object
```{r, message=FALSE, warning=FALSE}
library(phyloseq)
# building a phyloseq object from raw data
otu1 <- otu_table(counts, taxa_are_rows = TRUE)
sam1 <- sample_data(as.data.frame(pd))
tax1 <- tax_table(taxa_table)
PHYExp <- phyloseq(otu1, sam1, tax1)
```

# Testing each function
## `[` | Standard extraction operator. Works on `otu_table`, `sample_data`, and `taxonomyTable`
```{r}
# phyloseq
otu_table(PHYExp)[, 1] # phyloseq
assays(MBExp)[[1]][, 1] # MicrobiomeExperiment

sample_data(PHYExp)[, 1] # phyloseq
colData(MBExp)[, 1] # MicrobiomeExperiment

tax_table(PHYExp)[, 1] # phyloseq
rowData(MBExp)@hierarchy_tree[, 2] # MicrobiomeExperiment
```
## `access` | General slot accessor function for MicrobiomeExperiment-package
```{r}
# phyloseq
access(PHYExp, "otu_table", errorIfNULL = FALSE)
# MicrobiomeExperiment
slot(MBExp, "assays")
```
## `get_taxa` | Abundance values of all taxa in sample `i`
```{r}
# phyloseq
get_taxa(PHYExp, 1)
# MicrobiomeExperiment
assays(MBExp)[[1]][, 1]
```
## `get_sample` | Abundance values of taxa `i` for all samples
```{r}
# phyloseq
get_sample(PHYExp, 1)
# MicrobiomeExperiment
assays(MBExp)[[1]][1, ]
```
## `get_taxa_unique` | A unique vector of the observed taxa at a particular taxonomic rank
```{r}
# phyloseq
get_taxa_unique(PHYExp, taxonomic.rank = rank_names(PHYExp)[2])
# MicrobiomeExperiment
unique(as.data.frame(rowData(MBExp))[2])
```
## `get_variable` | An individual sample variable vector/factor
```{r}
# phyloseq
get_variable(PHYExp, varName = "b")
# MicrobiomeExperiment
colData(MBExp)[["b"]]
```
## `nsamples` | Get the number of samples described by an object
```{r}
# phyloseq
nsamples(PHYExp)
# MicrobiomeExperiment
ncol(assays(MBExp)[[1]])
```
## `ntaxa` | get the number of OTUs (taxa) described by an object
```{r}
# phyloseq
ntaxa(PHYExp)
# MicrobiomeExperiment
nrow(assays(MBExp)[[1]])
```
## `otu_table` | build or access `otu_table` objects

**NOTE: phyloseq does not extend SummarizedExperiment, so it does not have an "assays" slot, and therefore can only store one `otu_table` object**
```{r}
# phyloseq
otu_table(PHYExp) 
# MicrobiomeExperiment
# no equivalent function for build, but for accession:
assays(MBExp)[[1]]
```
## `rank_names` | Get the names of available taxonomic ranks
```{r}
# phyloseq
rank_names(PHYExp) # default behavior names ranks ta1, ta2, etc.
# MicrobiomeExperiment
colnames(rowData(MBExp))
```
## `sample_data` | build or access `sample_data` objects
```{r}
# phyloseq
sample_data(PHYExp)
# MicrobiomeExperiment
colData(MBExp)
```
## `sample_names` | The names of all samples
```{r}
# phyloseq
sample_names(PHYExp) # default behavior names samples sa1, sa2, etc.
# MicrobiomeExperiment
colData(MBExp)[["a"]]
```
## `taxa_names` | The names of all taxa
```{r}
# phyloseq
taxa_names(PHYExp) # default behavior names samples sp1, sp2, etc.
# MicrobiomeExperiment
splitAt(rowData(MBExp), selectedLevel = 8)
```
## `sample_sums` | The sum of the abundance values of each sample
```{r}
# phyloseq
sample_sums(PHYExp) # default behavior names samples sa1, sa2, etc.
# MicrobiomeExperiment
colSums(assays(MBExp)[[1]])
```
## `sample_variables` | The names of sample variables
```{r}
# phyloseq
sample_variables(PHYExp)
# MicrobiomeExperiment
colnames(colData(MBExp))
```
## `taxa_sums` | The sum of the abundance values of each taxa
```{r}
# phyloseq
taxa_sums(PHYExp)
# MicrobiomeExperiment
rowSums(assays(MBExp)[[1]])
```
## `taxa_are_rows` | `TRUE` if taxa are row indices in `otu_table`
```{r}
# phyloseq
taxa_are_rows(PHYExp)
# MicrobiomeExperiment
class(rowData(MBExp)) == "MicrobiomeFeatures" # if the rowData from the MBExp object is of class "MicrobiomeFeatures" then the row contains taxa information
```
## `tax_table` | A taxonomy table
```{r}
# phyloseq
tax_table(PHYExp)
# MicrobiomeExperiment
as.data.frame(rowData(MBExp)) # returns two additional columns: 'allFeatures' and 'otu_index'
```
## `phy_tree` | Access the tree contained in a MicrobiomeExperiment object
No corresponding accessor function in MicrobiomeExperiment or MicrobiomeFeatures
```{r}
phy_tree(PHYExp)
```

